---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
devtools::load_all()
```

# taxafmt

<!-- badges: start -->
<!-- badges: end -->

The goal of taxafmt is to format and parse taxonomic assignments, especially
for microbiome data sets.

## Installation

You can install the development version of taxafmt like so:

``` r
# install.packages("devtools")
devtools::install_github("kylebittinger/taxafmt")
```

## Example

The `taxafmt` package comes with a small data set of taxonomic assignments.

```{r example}
sprockett_taxa
```

Here, we'll show how to parse and format the assignments for use in plots or
statistical comparisons. Our first goal is to split up the taxa into ranks,
like family, genus, and species. The `split_lineage` function does this for us.

```{r}
split_lineage(sprockett_taxa$lineage)
```

That's okay, but we'd prefer to have the taxa lined up beside the Feature ID's.
We can use the `tidyverse` for that.

```{r message=FALSE}
library(tidyverse)
sprockett_taxa %>%
  mutate(split_lineage(lineage))
```

In some taxonomic systems like this one, the taxa are given a one-letter prefix
to indicate the rank. This is convenient, especially when we want to double
check that the correct taxon is in each column, but it's too ugly to use in
plots and reports. The funtion `remove_rank_prefix` will remove this prefix.

```{r}
sprockett_taxa %>%
  mutate(split_lineage(lineage)) %>%
  mutate(across(Kingdom:Species, remove_rank_prefix))
```

The species names are not quite right--they only contain the specific epithet,
which is the second word of the species name. The first word in the species
name is the genus. To generate real species names, we can use the function
`make_binomial_name`.

```{r}
sprockett_taxa %>%
  mutate(split_lineage(lineage)) %>%
  mutate(across(Kingdom:Species, remove_rank_prefix)) %>%
  mutate(Species = make_binomial_name(Genus, Species))
```


Finally, we'd like to create a shortened label for each taxonomic assignment,
something that's nice to use in a plot or report. The function `format_taxa`
does this. We give `format_taxa` a data frame of the taxa, and it generates
labels using the lowest-ranking taxon in the data frame (which should be on
the right hand side). If that taxon is `NA`, then we get a label containing
the lowest-rank taxon that's available, but tagged to indicate that it's not
classified all the way down.

Even though we worked to tidy up the species names, we con't really trust
species-level assignments for a data set like this. Therefore, we use the
`pick` function from `dplyr` to select the taxa from `Kingdom` to `Genus`
when making the labels.

```{r}
sprockett_taxa %>%
  mutate(split_lineage(lineage)) %>%
  mutate(across(Kingdom:Species, remove_rank_prefix)) %>%
  mutate(Species = make_binomial_name(Genus, Species)) %>%
  mutate(label = format_taxa(pick(Kingdom:Genus)))
```

Now, let's check out those labels.

```{r}
sprockett_taxa %>%
  mutate(split_lineage(lineage)) %>%
  mutate(across(Kingdom:Species, remove_rank_prefix)) %>%
  mutate(Species = make_binomial_name(Genus, Species)) %>%
  mutate(label = format_taxa(pick(Kingdom:Genus))) %>%
  select(feature_id, label)
```

